---
title: "GEOG 714 - Notes 2"
author: "Haoran Xu"
date: "Sep 16, 2024"
output:
  pdf_document:
    latex_engine: xelatex
header-includes:
  - \usepackage{listings}
  - \lstset{breaklines=true}  # 使代码块中的代码自动换行
---

## review of last course
```{r}
a <- c(2, 4, 3, 5, 7, 6)
print(length(a[1:3]))

print(order(c(30, 10, 20)))

j <- c(2, 3, 5, 4, 3, 2, 3, 6, 7)
k <- c(1, 4, 5, 2, 6, 7, 8, 9, 8)
m <- cbind(j, k)
m[,2]
```


data.frame() 
# object-attributes structure


```{r}
data.frame() 
# object-attributes structure
```

PCA --> look at Attributes
Clustering --> look at observations
Modelling --> look at all things


Shared Attribute <- Population
- parameters = 1.78m (average height of Canadian)  (it's unkownable, it's just statistic value out of samples) 
\pu often is measured as mean. (standard of the population)

sampling error 抽样误差
other types of errors: bias (equippment errors, method errors)
systematic error 系统误差/systematic bias

effect size - health

statistically significant does not mean "effect size" # 

p-value 有一些问题
p-value没有考虑任何practical significance (which it ignores effect size)

p-value is how likely you are to have found a particular set of observations if the null hypothesis were true.

null hypothesis: \mu_A = \mu_B

如果p-value特别小，说明inconsist with null hyposesis --> the lengths are different

sample size越大,p更容易显著？

需要注意：sample size很大，但是容易显著，但实际上effect很小（这叫p hacking)
